---
title: "Session-03-04 Sample Symportal run and downstream analyses"
format: gfm
editor: visual
---

# Objects, classes & variables

-   Everything in R is an **object**--data, function, etc.

-   The blueprint of an object is its **class**--description of the object.

-   A **variable** is an R object that contains an object.

```{r}
#| eval: false

1 # is an object,
class(1) # a numeric object or class numeric

'one' # is an object,
class('one') # a string object

'1' # what is this?
class('1') # a string object

a # why does this not show? # it has nothing attached to it
'a' # why is this okay?, it is designated as a string object

# Note that any alphabet including "." and "_" can be used as a variable name (as long as it is preceded by an alphabet)

a <- 1 # assigning an object to a variable
```

# Functions

An R function is an object containing statements that are run together in an order.

Can be a built-in (a base package \[no importation needed, included in R when you downloaded it\]) or user defined 🤓

Algebra days... 🍎\
f(x) = x \* x (the function of "f" at x)

For example, x = 3, the result will be 9. We know that this function squares any numeric object.

```{r}
square_a_stuff <- 
  function(x){
			squared_stuff = x * x
			return(squared_stuff)
			}
```

How do we use a function?

bash scripting: `conda` **activate** *qiime2.2023.7*

```{r}
square_a_stuff(8)
```

# Import packages

Before starting, it is important to import the important packages that will be used in a run. 📖

For this tutorial, the main packages are: `tidyverse` and `vegan`.

Note that there are many more packages that can be used but for the purpose of this workshop... we'll just be using these two. 🤓

First, you need to download the packages.

```{r}
#| eval: false
install.packages(c('tidyverse, vegan'))
```

Then import/call them.

```{r}
library('tidyverse',
        'vegan',
        'ape')
```

# Import and clean data

## Metadata

Import.

```{r}
metadata <- 
  read_csv('../data/Session-03-04-metadata.csv')
```

Clean.

```{r}
metadata1 <- 
  metadata %>% 
  filter(coi_validated_species == 'Ts'| coi_validated_species == 'Tg') # Ts and Tg are wanted
```

## Symportal ITS2 sequences results

Import.

```{r}
its2seq <- 
  read_tsv('../data/Session-03-04-symportal-its2seq.txt')

# Check dataframe
head(its2seq)
tail(its2seq)
```

Clean.

```{r}
its2seq1 <- 
  its2seq %>% 
  slice(1:190)

colnames(its2seq1) # Large number of
```

Make it cleaner.

```{r}
its2seq2 <- 
  its2seq1 %>% 
  select(2,40:500) %>% 
  filter(sample_name %in% pull(metadata1, transformed))
```

## Merge metadata and ITS2 sequences

Merge metadata and ITS2 sequences and clean. 🧹

```{r}
its2seq_metadata <- 
  metadata1 %>%
  select(!sample_name) %>% 
  rename(sample_name = transformed) %>% 
  left_join(its2seq2, by = 'sample_name') %>% 
  select(!c(1:6)) %>% 
  select(!c(2,4,6,7))
```

# Data analysis

## Abundance plots

### Genus

Clean data.

```{r}
gen_metadata <- 
  its2seq_metadata %>% 
  rowwise() %>% 
  mutate(A = sum(c_across(contains('A', ignore.case = F))),
         C = sum(c_across(contains('C', ignore.case = F))),
         D = sum(c_across(contains('D', ignore.case = F))),
         G = sum(c_across(contains('G', ignore.case = F)))) %>% 
  select(c(1:3, 465:468))

gen_metadata1 <- 
  gen_metadata %>% 
  pivot_longer(cols = c(4:7),
               names_to = 'Genus',
               values_to = 'Count') %>% 
  pivot_longer(cols = c(2:3),
               names_to = 'facet_param',
               values_to = 'x')
```

Make plot.

```{r}
gen_metadata1_plot <- 
  gen_metadata1 %>% 
  mutate(x = str_replace(x, 'Tridacna', 'T.')) %>% 
  ggplot(aes(x = x,
             y = Count,
             fill = Genus)) +
  geom_bar(position="fill", 
           stat="identity") +
  facet_grid(cols = vars(facet_param),
             scales = 'free_x',
             space = 'free_x') +
  labs(y = 'Relative abundance (%)') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank())

ggsave(gen_metadata1_plot,
       filename = '../results/gen_metadata1_plot.png',
       width = 300,
       height = 150, 
       units = "mm",
       type = 'cairo',
       device = 'png')
```

### ITS2 sequence

clean data.

```{r}
its2seq_metadata1 <- 
  its2seq_metadata %>% 
  pivot_longer(cols = !c(1:3),
               names_to = 'ITS2',
               values_to = 'Count') %>% 
  pivot_longer(cols = c(2:3),
               names_to = 'facet_param',
               values_to = 'x')
```

Make plot.

```{r}
its2seq_metadata1_plot <- 
  its2seq_metadata1 %>% 
  mutate(x = str_replace(x, 'Tridacna', 'T.')) %>% 
  ggplot(aes(x = x,
             y = Count,
             fill = ITS2)) +
  geom_bar(position="fill", 
           stat="identity") +
  facet_grid(cols = vars(facet_param),
             scales = 'free_x',
             space = 'free_x') +
  labs(y = 'Relative abundance (%)') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank())
```

Get top 10.

```{r}
its2seq_metadata2 <- 
  its2seq_metadata %>% 
  select(!c(2:3)) %>% 
  pivot_longer(cols = !1,
               names_to = 'ITS2',
               values_to = 'Count') %>% 
  group_by(ITS2) %>% 
  summarise(`Total count` = sum(Count)) %>% 
  arrange(desc(`Total count`)) %>% 
  mutate(`Relative abundance (%)` = round(`Total count`/6750449 * 100,2))

write_csv(its2seq_metadata2,
          '../results/its2seq_metadata2_relativeabundance.csv')
```

Make a plot.

```{r}
its2seq_metadata1_plot <- 
  its2seq_metadata1 %>% 
  mutate(x = str_replace(x, 'Tridacna', 'T.')) %>% 
  filter(ITS2 %in% pull(slice(its2seq_metadata2, 1:10), ITS2)) %>% 
  ggplot(aes(x = x,
             y = Count,
             fill = ITS2)) +
  geom_bar(position="fill", 
           stat="identity") +
  facet_grid(cols = vars(facet_param),
             scales = 'free_x',
             space = 'free_x') +
  labs(y = 'Relative abundance (%)') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank())

ggsave(its2seq_metadata1_plot,
       filename = '../results/its2seq_metadata1_plot.png',
       width = 300,
       height = 150, 
       units = "mm",
       type = 'cairo',
       device = 'png')
```

Get relative abundance per species and site.

```{r}
its2seq_metadata1_relativeabundance <- 
  its2seq_metadata1 %>% 
  group_by(x, ITS2) %>% 
  summarise(`Total count` = sum(Count)) %>% 
  mutate(group = case_when(str_detect(x, 'Tridacna') ~ 'GC species',
                           .default = 'Site'))

its2seq_metadata1_relativeabundance_site <- 
  its2seq_metadata1_relativeabundance %>% 
  filter(group == 'Site')  %>% 
  mutate('Relative abundance' = `Total count`/sum(`Total count`)*100)

write_csv(its2seq_metadata1_relativeabundance_site,
          '../results/its2seq_metadata2_relativeabundance_site.csv')

###

its2seq_metadata1_relativeabundance_gc <- 
  its2seq_metadata1_relativeabundance %>% 
  filter(group == 'GC species')  %>% 
  mutate('Relative abundance' = `Total count`/sum(`Total count`)*100)

write_csv(its2seq_metadata1_relativeabundance_gc,
          '../results/its2seq_metadata2_relativeabundance_gc.csv')
```

## Beta diversity

### Make matrices

Make a metadata matrix.

Note: most vegan packages use matrices as input, not matrix, error! ⚠️

```{r}
metadata1_matrix <- 
  metadata1 %>% 
  select(!sample_name) %>% 
  rename(sample_name = transformed) %>% 
  select(sample_name, comple_species, region) %>% 
  arrange(desc(sample_name)) %>% 
  mutate(sample_name2 = sample_name) %>% 
  column_to_rownames('sample_name2')
```

Make an abundance table matrix.

```{r}
its2seq2_matrix <- 
  its2seq_metadata %>% 
  select(!c(2,3)) %>% 
  arrange(desc(sample_name)) %>% 
  column_to_rownames('sample_name')
```

### Compute Bray-Curtis

Compute Bray-Curtis

```{r}
its2seq2_matrix_bc <- 
  vegan::vegdist(its2seq2_matrix)

write.csv(its2seq2_matrix_bc %>% as.matrix(),
          '../results/its2seq2_matrix_bc.csv')
```

### PCoA

Bray-Curtis ➡️ PCoA.

```{r}
its2seq2_matrix_bc_pcoa <- 
  ape::pcoa(its2seq2_matrix_bc)

its2seq2_matrix_bc_pcoa1 <- 
  ape::pcoa(its2seq2_matrix_bc) %>% 
  .$vectors %>% 
  as_data_frame() %>% 
  rownames_to_column('bc_name')

metadata1_matrix1 <- 
  metadata1_matrix %>% 
  mutate(bc_name = 1:n())

its2seq2_pcoa_meta <- 
  its2seq2_matrix_bc_pcoa1 %>% 
  type_convert() %>% 
  left_join(metadata1_matrix1,
            by = 'bc_name') %>% 
  relocate(36:38)
```

Plot PCoA for species comparison.

```{r}
its2seq2_pcoa_meta_plot_species <- 
  its2seq2_pcoa_meta %>% 
  ggplot(aes(x = Axis.1,
             y = Axis.2,
             color = comple_species)) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'gray') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'gray') +
  geom_point(size=4) +
  stat_ellipse() +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Plot PCoA for site.

```{r}
its2seq2_pcoa_meta_plot_site <- 
  its2seq2_pcoa_meta %>% 
  ggplot(aes(x = Axis.1,
             y = Axis.2,
             color = region)) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'gray') +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'gray') +
  geom_point(size=4) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

### PERMANOVA

```{r}
its2seq2_pcoa_permanova <- 
  vegan::adonis2(its2seq2_matrix_bc ~ `comple_species` * `region`, 
          data = metadata1_matrix1, 
          permutations = 9999)

write_csv(its2seq2_pcoa_permanova %>% rownames_to_column('Parameters'),
          '../results/its2seq2_pcoa_permanova.csv')
```

### PERMDISP

```{r}
its2seq2_pcoa_permdisp_gc <- 
  vegan::betadisper(its2seq2_matrix_bc,
             metadata1_matrix1 %>% pull(comple_species)) %>% 
  anova() %>% 
  rownames_to_column('Parameters') %>% 
  as_tibble() %>% 
  drop_na() %>% 
  mutate('Group comparison' = 'Giant clam species') %>% 
  relocate('Group comparison')

write_csv(its2seq2_pcoa_permdisp_gc,
          '../results/its2seq2_pcoa_permdisp_gc.csv')
  
its2seq2_pcoa_permdisp_site <- 
  vegan::betadisper(its2seq2_matrix_bc,
             metadata1_matrix1 %>% pull(region)) %>% 
  anova() %>% 
  rownames_to_column('Parameters') %>% 
  as_tibble() %>% 
  drop_na() %>% 
  mutate('Group comparison' = 'Site') %>% 
  relocate('Group comparison')

write_csv(its2seq2_pcoa_permdisp_gc,
          '../results/its2seq2_pcoa_permdisp_site.csv')
```

## Alpha diversity

Compute alpha diversity.

```{r}
its2seq_metadata_alpha <- 
  its2seq_metadata %>% 
  pivot_longer(cols = !c(1:3),
               names_to = 'ITS2',
               values_to = 'Count') %>% 
  group_by(sample_name) %>% 
  summarise(observed_its2seqs = vegan::specnumber(Count),
            shannon = vegan::diversity(Count)) %>% 
  left_join(metadata1_matrix,
            by = 'sample_name')
```

### Boxplots

Clean data.

```{r}
its2seq_metadata_alpha_plot <- 
  its2seq_metadata_alpha %>% 
  mutate(comple_species = str_replace(comple_species, 'Tridacna', 'T.')) %>% 
  pivot_longer(cols = c(2,3),
               names_to = 'alpha_metric',
               values_to = 'alpha_value') %>% 
  pivot_longer(cols = c(comple_species, region),
               names_to = 'facet_parameter',
               values_to = 'group') %>% 
  ggplot(aes(x = group,
             y = alpha_value)) +
  geom_violin(trim = F,
              color = NA,
              alpha = 0.5,
              fill = 'black') +
  geom_boxplot(fill = 'black',
               width = 0.04,
               outlier.shape = NA) + 
  stat_summary(geom = 'point',
               fun = median,
               shape = 21,
               size = 2,
               fill = 'white') +
  facet_grid(rows = vars(alpha_metric),
             cols = vars(facet_parameter),
             scales = 'free',
             space = 'free_x') +
  theme_bw() +
  theme(panel.grid = element_blank())

ggsave(its2seq_metadata_alpha_plot,
       filename = '../results/its2seq_metadata_alpha_plot.png',
       width = 300,
       height = 150, 
       units = "mm",
       type = 'cairo',
       device = 'png')
```

### Kruskal-Wallis (non-parametric ANOVA)

Do stats.

```{r}
its2seq_metadata_alpha_site_kw <- 
  its2seq_metadata_alpha %>% 
  kruskal.test(observed_its2seqs ~ region, data = .)

write_csv(its2seq_metadata_alpha_site_kw %>% unclass() %>% as_tibble(),
          '../results/its2seq_metadata_alpha_site_kw.csv')

its2seq_metadata_alpha_gc_kw <- 
  its2seq_metadata_alpha %>% 
  kruskal.test(observed_its2seqs ~ comple_species, data = .)

write_csv(its2seq_metadata_alpha_gc_kw %>% unclass() %>% as_tibble(),
          '../results/its2seq_metadata_alpha_gc_kw.csv')
```
